Domain-agnostic Outlier Ranking Algorithms (DORA)---A Configurable Pipeline for Facilitating Outlier Detection in Scientific Datasets

### 2021-09-19 publications/iaai21/dora_outlier_detection_iaai.tex
Trim text, fill in abstract/intro todos. #42

\begin{abstract}
Automatic detection of out-of-distribution (OOD) samples or features is
universally needed when working with scientific datasets. For example,
OOD detection methods can be used for cleaning datasets or flagging
novel samples in order to guide instrument acquisition or scientific analysis.
We present Domain-agnostic Outlier Ranking Algorithms (DORA) to provide
a configurable pipeline that makes it easy for scientists to apply and evaluate
outlier detection methods in a variety of domains. DORA allows users to 
configure outlier detection experiments by specifying the location of their 
dataset(s), the input data type, feature extraction methods, and which 
algorithms should be applied. DORA supports image, raster, time series, 
or feature vector input data types and outlier detection methods that include
 Isolation Forest, Discovery via Eigenbasis Modeling of Uninteresting Data 
 (DEMUD), principal component analysis (PCA), Reed-Xiaoli (RX) detector, 
 Local RX, negative sampling, and probabilistic autoencoder. Given a set of
  data samples, each algorithm assigns an outlier score to each sample. DORA
   provides results organization and visualization modules to help users 
   process the results, including sorting samples by outlier score, evaluating 
   outlier recall for a set of known/labeled outliers, clustering groups of 
   similar outliers together, and causal graphs. To demonstrate the 
   cross-domain applicability of DORA, we conducted outlier detection
    experiments for three real-world use cases: 1) cleaning noisy ground-truth 
    datasets (Earth Science), 
     2) flagging non-astrophysical objects in the Dark Energy Survey 
     (astrophysics), and 3) novelty-guided onboard targeting of instruments
      for Mars rovers (planetary science). \todo{[We showed that ...]}
\end{abstract}

@@

   When working with science datasets, OOD detection can be used for 
   cleaning datasets, e.g., flagging ground-truth labels with GPS or human
    entry error or identifying wrongly categorized objects in a catalog. 
    It could also be used for discovery, e.g., to flag novel samples in order 
    to guide instrument acquisition or scientific analysis. Another application
   is the detection of rare objects that are known to exist but the known
   examples are too few to create a large enough labeled dataset for 
   supervised classification algorithms. 
  Despite wide differences in applications, data types, and dimensionality,
 the same underlying machine learning algorithms can be employed across 
 all of these domains. A challenge for applying them however is that domain
 scientists do not always have the programming or machine learning background
 to apply the algorithms themselves using existing tools. Given the widespread 
 applicability and transferability of OOD methods, the scientific community 
 would benefit from a tool that made it easy for them to apply popular outlier
 detection algorithms to their science datasets. We created DORA (Domain-
 agnostic Outlier Ranking Algorithms) to provide a tool for applying outlier 
 detection algorithms to a variety of scientific data sets with minimal coding
 required. Users need only to specify details for their data/application 
 including the data type, location, and algorithms to run in an experiment
 configuration file. DORA supports image, raster, time series, 
or feature vector input data types and outlier detection methods that include
 Isolation Forest, Discovery via Eigenbasis Modeling of Uninteresting Data 
 (DEMUD)~\citep{wagstaff:demud13}, principal component analysis (PCA),
   Reed-Xiaoli (RX) detector,  
 Local RX, negative sampling~\cite{sipple:neg-sampling20}, and probabilistic 
 autoencoder. Given a set of data samples, each algorithm assigns an outlier 
 score to each sample. DORA provides results organization and visualization 
 modules to help users process the results, including sorting samples by outlier 
 score, evaluating outlier recall for a set of known/labeled outliers, clustering 
 groups of similar outliers together, and causal graphs. To demonstrate the 
 cross-domain applicability of DORA, we conducted outlier detection experiments 
 for three real-world use cases: 1) cleaning noisy ground-truth datasets (Earth 
 Science), 2) flagging non-astrophysical objects in the Dark Energy Survey 
 (astrophysics), and 3) novelty-guided onboard targeting of instruments for Mars 
 rovers (planetary science). \todo{[We showed that ...]}

@@

\subsubsection{Benchmark: MNIST and Fashion-MNIST.}
We used MNIST and F-MNIST to demonstrate the capabilities of DORA with a
traditional benchmark dataset. We used 60,000 images from 
F-MNIST as the training set and a test set of 1,000 images from MNIST (outliers)
 and 1,000 from F-MNIST (inliers). 
 %Well-discriminating algorithms should assign 
%higher outlier scores to samples from MNIST than F-MNIST.

>>>>

\begin{abstract}
Automatic detection of outliers is
universally needed when working with scientific datasets, e.g., for cleaning 
datasets or flagging novel samples to guide instrument acquisition or
 scientific analysis.
We present Domain-agnostic Outlier Ranking Algorithms (DORA),
a configurable pipeline that facilitates application and evaluation of
outlier detection methods in a variety of domains. DORA allows users to 
configure experiments by specifying the location of their 
dataset(s), the input data type, feature extraction methods, and which 
algorithms should be applied. DORA supports image, raster, time series, 
or feature vector input data types and outlier detection methods that include
 Isolation Forest, DEMUD, PCA, RX detector, 
 Local RX, negative sampling, and probabilistic autoencoder. Each algorithm assigns an outlier score to each data sample. DORA
   provides results interpretation modules to help users 
   process the results, including sorting samples by outlier score, evaluating 
   the fraction of known outliers in $n$ selections, clustering groups of 
   similar outliers together, and web visualization. We demonstrated how DORA
   facilitates application, evaluation, and interpretation of outlier detection
   methods by performing
    experiments for three real-world datasets from Earth science, planetary
  science, and astrophysics, as well as one benchmark dataset 
  (MNIST/Fashion-MNIST). We found that no single algorithm
      performed best across all datasets, underscoring the need for a tool 
      that enables comparison of multiple algorithms.
\end{abstract}

@@

   When working with science datasets, OOD detection can be used for 
   cleaning datasets, e.g., flagging ground-truth labels with GPS or human
    entry error or identifying wrongly categorized objects in a catalog. 
    It could also be used for discovery, e.g., to flag novel samples in order 
    to guide instrument acquisition or scientific analysis. Another application
   is the detection of rare objects that are known to exist but the known
   examples are too few to create a large enough labeled dataset for 
   supervised classification algorithms. 
  Despite wide differences in applications, data types, and dimensionality,
 the same underlying machine learning algorithms can be employed across 
 all of these domains. A challenge for applying them however is that domain
 scientists do not always have the programming or machine learning background
 to apply the algorithms themselves using existing tools. Given the widespread 
 applicability and transferability of OOD methods, the scientific community 
 would benefit from a tool that made it easy for them to apply popular outlier
 detection algorithms to their science datasets. We created DORA 
 (Domain-agnostic Outlier Ranking Algorithms) 
 to provide a tool for applying outlier 
 detection algorithms to a variety of scientific data sets with minimal coding
 required. Users need only to specify details for their data/application 
 including the data type, location, and algorithms to run in an experiment
 configuration file. DORA supports image, raster, time series, 
or feature vector input data types and outlier detection methods that include
 Isolation Forest, Discovery via Eigenbasis Modeling of Uninteresting Data 
 (DEMUD)~\citep{wagstaff:demud13}, principal component analysis (PCA),
   Reed-Xiaoli (RX) detector,  
 Local RX, negative sampling~\cite{sipple:neg-sampling20}, and probabilistic 
 autoencoder (PAE). 
 Each algorithm assigns an outlier 
 score to each sample in a given dataset. 
 DORA provides results organization and visualization 
 modules to help users process the results, including sorting samples by outlier 
 score, evaluating outlier recall for a set of known/labeled outliers, clustering 
 groups of similar outliers together, and web visualization. 
 We demonstrated how DORA
   facilitates application, evaluation, and interpretation of outlier detection
   methods by performing
    experiments for three real-world datasets from Earth science, planetary
  science, and astrophysics, as well as one benchmark dataset 
  (MNIST/Fashion-MNIST). We found that no single algorithm
      performed best across all datasets, underscoring the need for a tool 
      that enables comparison of multiple algorithms.

@@

\subsubsection{Benchmark: MNIST and Fashion-MNIST.}
We used MNIST and F-MNIST to demonstrate DORA with a
traditional benchmark dataset. We used 60,000 images from 
F-MNIST as the training set and a test set of 1,000 images each from MNIST (outliers) and F-MNIST (inliers). 
 %Well-discriminating algorithms should assign 
%higher outlier scores to samples from MNIST than F-MNIST.
